\chapter{Background and Terminology}
\label{chp:background}

This chapter details event sourcing and related terminology. We provide an 
overview on benefits and constraints of event-sourced systems and examine 
in which cases event sourcing can be applied beneficially as a style of 
architecture. 
Furthermore, we describe Command Query Responsibility Segregation (CQRS), 
a complementary architectural style for event-sourced systems, and illustrate 
which benefits can be gained from applying it.

\input{chapters/es-intro}
\input{chapters/cqs-cqrs}

\section{Summary}
In this chapter, we detailed the concept of event sourcing. We laid out the
terminology used in this thesis and described the Command Query Responsibility 
Segregation (CQRS) pattern, a common style of architecture for event-sourced 
systems. Furthermore, we examined the symbiotic relationship formed by CQRS 
and event sourcing.

Although we described some of the benefits which event-sourced systems can 
provide, it is important to note that event sourcing is seldom applied to an 
overall system. Indeed, an often found recommendation is to apply this style 
of architecture only within a clearly defined context. The same recommendation 
is often found for CQRS, since the command/query model segregation comes with 
the implication of higher complexity and often an eventually consistent 
behavior. Thus, applying CQRS as a style of architecture is only beneficial 
when the system has requirements for CQRS properties, such as individual 
optimization of the read and write model.
